
# Supervised and unsupervised learning

## Introduction

The methods in **machine learning** can be broadly divided into **supervised** and
**unsupervised** learning methods.  

In **supervised learning**, we have an outcome variable $Y$ (also called output,
response, dependent variable) and a set of predictors $\mathbf{X}$ (also called 
features, dependent variables, covariates). The objective is to estimate 
the function $f(\mathbf{X})$ that connects $Y$ and $\mathbf{X}$. For example,

\[
Y = f(\mathbf{X}) + \varepsilon
\]
 
(if $Y$ is categorical we generally consider the association of $\mathbf{X}$ and
the $Pr(Y=k)$)

Having a dataset with observations of $Y$ and $X$ (called training set), we can
use these data to find an estimate $\hat f(\mathbf{X})$ according to some 
optimisation principle.

If we specify a functional for $f(\mathbf{X})$, such as the linear regression 
model:$f(\mathbf{X}) = \beta_0 + \beta_1X_1 + ... + \beta_p X_p$, we say that 
we are using a *parametric* method. If instead we allow the data to estimate
the functional form of $f(\mathbf{X})$, such as k-nearest neighbourhood 
regression, we call the method *non-parametric*.

In **unsupervised learning** , the goal is to model the 
underlying structure or distribution in the data without having the outcome $Y$.
In other words, we want to analyse how the features $\mathbf{X}$ are *clustered*
or *associated*.  Examples of these methods are *principle components analysis*
and *k-means clustering*.


## Readings

Read the following chapters of *An introduction to statistical learning*: 

* 2.1 [What is statistical learning](http://faculty.marshall.usc.edu/gareth-james/ISL/ISLR%20Seventh%20Printing.pdf#page=28)

If you are using R for the first time, I encourage you to complete the section

* 2.3 [Introduction to R](http://faculty.marshall.usc.edu/gareth-james/ISL/ISLR%20Seventh%20Printing.pdf#page=5)



## R review

### Task 1 - Read a dataset and create a new variable {-}

Read the [bmd.csv](https://www.dropbox.com/s/7wjsfdaf0wt2kg2/bmd.csv?dl=1) 
dataset in R.  You can download and read the data or use the path *https://www.dropbox.com/s/7wjsfdaf0wt2kg2/bmd.csv?dl=1*

```{r message = FALSE}
bmd.data <- read.csv("https://www.dropbox.com/s/7wjsfdaf0wt2kg2/bmd.csv?dl=1")
```

If you have downloaded the file, you need to substitute the *url* above by 
the path to file in your computer, e.g., "c:\\my files\\bmd.csv"

The variable **id** is the identification of the subjects in the dataset, and
the variable **age** is their age. What is the age of the subject id=197?

`bmd.data$age` is the vector with the ages of all the subjects. Within that 
vector we need to find the one where `bmd.data$id == 197`.  Notice that double
equal sign "==" is a logic statement. 

If we write `bmd.data$id = 197` we are 
assigning the value 197 to the variable **id** and every subject in the dataset
gets a new value 197 for that variable.

```{r message = FALSE}

bmd.data$age[bmd.data$id == 197]
#or equivalent
bmd.data[bmd.data$id == 197, "age"]
#or 
bmd.data[bmd.data$id == 197, 2]  #age is the second variable
```

Let's now create a new variable **age.months** which is age in months (the 
original is in years).

```{r message = FALSE}
# we just need to multiply the original by 12
bmd.data$age.months <- bmd.data$age * 12

# we can use = instead of <- 
bmd.data$age.months = bmd.data$age * 12
```

Finally, let's recode the variable **age** into age categories <=50, (50,60],
(60,70] and >60.

```{r message = FALSE}
# we can use the function cut() to 
#create the categories
bmd.data$age.cat <- cut(bmd.data$age, 
                        c(0,50,60,70, 100))

# alternative 
bmd.data$age.cat2[bmd.data$age<50]                    <- 1
bmd.data$age.cat2[bmd.data$age>=50 & bmd.data$age<60] <- 2
bmd.data$age.cat2[bmd.data$age>=60 & bmd.data$age<70] <- 3
bmd.data$age.cat2[bmd.data$age>=70]                   <- 4

#frequencies of the categories
table(bmd.data$age.cat)

```

**TRY IT YOURSELF:**

1) Get the ages for all the male subjects, i.e., `bmd.data$sex == "M"`.

<details><summary>See the solution code</summary>

```{r , results="hide"}
bmd.data$age[bmd.data$sex == "M"]  
```
</details><p><p>

2) What is the length of the vector above? Or, in other words, how many subjects 
are male?

<details><summary>See the solution code</summary>

```{r , results="hide"}
length(bmd.data$age[bmd.data$sex == "M"])

#or
table(bmd.data$sex)
```
</details><p><p>

3) Using the variables **weight_kg** and **height_cm**, compute the body mass
index? 
$\frac{\text{weight in Kg}}{(\text{heigh in m})^2}$

<details><summary>See the solution code</summary>

```{r , results="hide"}
bmd.data$bmi <- bmd.data$weight_kg /(bmd.data$height_cm/100)**2
```
</details><p><p>


### Task 2 - Histogram {-}

There are several packages specific to graphs
`ggplot2`, `lattice`, `plotly`, `highcharter`, `sunburstR`, `dygraphs`, `rgl`,…
The [R Graph Gallery](https://www.r-graph-gallery.com/index.html) is an excellent 
learning resource for complex plots and visualisation

We will start with basic functions. Let’s produce the histogram of the variable
**bmi** created in task 1

```{r hisbmi , }
hist(bmd.data$bmi)
```

Let’s edit the title and x-axis label

```{r hisbmi2 , }
	hist(bmd.data$bmi, 
	     main = "My 1st histogram", 	
	     xlab = "Body Mass Index")
```

And we can also plot the distribution for males and females, separately

```{r hisbmi3 , }

#histogram for males
hist(bmd.data$bmi[bmd.data$sex=="M" ],
     breaks = 10,                  # number of cutoff for the bars
     main= "My 1st histogram",  
     xlab= "Body Mass Index",
     col=rgb(0,0,1,.5))           # color of the bars

#histogram for females
hist(bmd.data$bmi[bmd.data$sex=="F" ],
      breaks = 10,
     main= "My 1st histogram",
     xlab= "Body Mass Index", 
     col=rgb(1,0,0,.5),
     add=T)                      # superimposes the histograms

# Add legend
legend("topright",  
       legend=c("Males","Females"), 
       col=c(rgb(0,0,1,0.5),  rgb(1,0,0,0.5)), 
       pt.cex=2, pch=15 )
```


### Task 3 - Scatter and boxplot {-}

To see the relation between bone mineral density (**BMD**) and **age** we 
can produce a scatter plot:

```{r scatage , }
# BMD 
plot(bmd.data$age,bmd.data$bmd)
```

Or a boxplot between bone mineral density (**BMD**) by categories of age created
in task 1

```{r bpage , }
# BMD 
boxplot(bmd.data$bmd ~ bmd.data$age.cat)
```

And with a bit more work:

```{r bpage2 , }
# BMD 
boxplot(bmd.data$bmd ~ bmd.data$age.cat, 
        main = "A nice boxplot (at least for a colorblind)",
        xlab = "Age categories",
        ylab = "Bone Mineral Density",
        col  = terrain.colors(4) ) 

# Add data points
mylevels         <- levels(bmd.data$age.cat) 
levelProportions <- summary(bmd.data$age.cat)/nrow(bmd.data) 

for(i in 1:length(mylevels)){ 
  thislevel  <- mylevels[i] 
  thisvalues <- bmd.data[bmd.data$age.cat == thislevel, "bmd"] 
  #take the x-axis indices and add a jitter, 
  #proportional to the N in each level 
  myjitter   <- jitter(rep(i, length(thisvalues)), 				
                       amount=levelProportions[i]/2) 
  points(myjitter, thisvalues, 
	pch=20, col=rgb(0,0,0,.9)) 
}
```


**TRY IT YOURSELF:**

1) Plot a scatter for the relation of **bmi** and **bmd**.

<details><summary>See the solution code</summary>

```{r xx1, results="hide"}
plot(bmd.data$bmi,bmd.data$bmd)
```
</details><p><p>

2) Plot a scatter for the relation of **bmi** and **bmd** with the dots
painted by the variable **sex**.

<details><summary>See the solution code</summary>

```{r xx2, results="hide"}
plot(bmd.data$bmi,bmd.data$bmd, 
     pch=20,
     col= ifelse(bmd.data$sex=="M",  #blue for males
                 "blue",             #red for females
                 "red")
     )

# Add legend
legend("topright",  
       legend=c("Males","Females"), 
       col=c("blue",  "red"), 
       pt.cex=3, pch=20 )
```

</details><p><p>




